A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-27 main.tex
Multiple automata are now a section

\section{Parikh Images from Products of Automata}\label{sec:multiple}

We now generalise our calculus to natively work with intersections of
regular languages, or equivalently products of automata. For this
extension, we change the main predicate~$\Image$ to be indexed by a
vector of automata~$\Tuple{\Automaton_1, \ldots, \Automaton_k}$.  For
simplicity, we assume that the sets of states of the $k$ automata (and
therefore also the transition sets) are pairwise disjoint. 
\begin{definition}\label{def:multiple}
  Suppose $\Automaton_1, \ldots, \Automaton_k$ are automata,
  $\Map : \Alphabet^* \to \Monoid$ is a map to a commutative monoid
  $\Monoid$,
  $\Filter: \bigcup_{i=1}^k \Transitions_{\Automaton_i} \to \Naturals$
  is a transition selection function, and
  $\MonoidElement \in \Monoid$. The predicate
  $\ImagePredicate{\Tuple{\Automaton_1, \ldots,
      \Automaton_k}}{\Map}{\Filter}{\MonoidElement}$ is true exactly
  when there are accepting paths~$\Path_1, \ldots, \Path_k$ of the
  respective automata, such that for each $i \in \{1, \ldots, k\}$ and
  $\Transition \in \Transitions_{\Automaton_i}$ it holds that
  \begin{itemize}
  \item the multiplicity of $\Transition$ on $\Path_i$ is consistent with
    $\Filter$, that is,
    $\Filter(\Transition) = \TransitionCount(\Transition, \Path_i)$,
  \item the automata all accept the same
    word~$\WordOf(\Path_i) = \WordOf(\Path_1)$, and
  \item the accepted word is mapped to
    $\MonoidElement = \Map(\WordOf(\Path_i)) =
    \Map(\WordOf(\Path_1))$.
  \end{itemize}
\end{definition}

  \begin{table}[t]
    \begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
      Name & \text{Rule} & Side conditions\\
      \midrule

      % EXPAND
      \ExpandM & 
      \inferrule
      {
       {\begin{array}{c}
           \left\{
           \FlowEq(\Automaton_i, \Filter),~~
           \Connected(\Automaton_i, \Filter),~~
           \MonoidElement = \sum_{\Transition \in \Transitions_{\Automaton_i}} \Filter(\Transition) \cdot \Map(\Transition)
           \right\}_{i=1}^k,
           \\[4ex]
          \ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, ~
          \SomeInequalities, \SomeClause
        \end{array}}
        }
      {\ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause} & 
      None \\[5ex]
      \Materialise &
      \inferrule
      {{
         \begin{array}{c}
           \FlowEq(\Automaton', \Filter'),~
           \Connected(\Automaton', \Filter'),~
           \BindingSum(\Automaton_i, \Automaton_j, \Filter')
           \\\ImagePredicate{
                     \Tuple{\Automaton_1, \Automaton_{i-1}, \Automaton_{i+1}, \ldots,
                     \Automaton_{j-1}, \Automaton_{j+1}, \ldots, \Automaton_k,
                     \Automaton'},~ \SomeInequalities, \SomeClause
           }{\Map}{\Filter'}{\MonoidElement}
           \end{array}}
           }
      {\ImagePredicate{\Tuple{\Automaton_1, \ldots, \Automaton_i, \ldots, \Automaton_j, \ldots, \Automaton_k}}{\Map}{\Filter}{\MonoidElement},~ \SomeInequalities, \SomeClause} &
      $\begin{array}{@{}l@{}}
         \Filter' = \Extend(\Automaton_i, \Automaton_j,\Filter),\\
         1 \leq i < j \leq k, \\
      \Automaton' = \Automaton_i \times \Automaton_j,\\
      \end{array}$ \\
      \bottomrule
      \end{tabular}
      \caption{Additional derivation rules for products of arbitrarily many automata.}\label{tbl:rules:multi}
    \end{table}

    For the calculus (\cref{tbl:rules:multi}), we first extend
    $\Expand$ to generate flow equations and instances of $\Connected$
    for each automaton, resulting in a new rule $\ExpandM$.  Unlike
    \Expand{}, \ExpandM{} does not remove the $\Image$~predicate,
    since it is needed to keep track of the currently considered
    partial products.

The rule~$\Materialise$ introduces the product of two individual
automata~$\Automaton_i, \Automaton_j$; this step eliminates
$\Automaton_i, \Automaton_j$ as index of the $\Image$ predicate, and
instead adds the product~$\Automaton_i \times \Automaton_j$, while
also introducing the flow equations and the $\Connected$ predicate.

The rule~$\Materialise$ also has to connect the newly introduced
product~$\Automaton_i \times \Automaton_j$ to the previous automata
$\Automaton_i, \Automaton_j$. This is done by extending the
selection function~$\Filter$ to $\Filter'$, mapping the transitions of the product
to fresh variables~$x_{\Transition}$:
\begin{equation*}
  \Extend(\Automaton_i, \Automaton_j,\sigma)(t)
  ~~=~~
  \begin{cases}
    x_{\Transition} & t \in \Transitions_{\Automaton_i \times \Automaton_j}
    \\
    \Filter(t) & \text{otherwise}
  \end{cases}
\end{equation*}

The multiplicity of transitions in the product then has to be related
to the multiplicities in the individual automata, modelled using the
$\BindingSum$ predicate. The predicate expresses that the multiplicity
of a transition~$\Transition \in \Transitions_{\Automaton_i}$ in
$\Automaton_i$ has to coincide with the sum of the multiplicities of
transitions in $\Automaton_i \times \Automaton_j$ derived from $t$,
and similarly for $\Automaton_j$:
%
  $$
  \BindingSum(\Automaton_i, \Automaton_j, \Filter') ~~=~~
  \begin{aligned}
  & \Set{ 
    \Filter'(\Transition)  =  \sum\limits_{\Transition' = \FromLabelTo{\Tuple{\State, \State_R}}{\Label}{\Tuple{\State', \State_R}}} \Filter'(\Transition')
  \SuchThat \Transition = \FromLabelTo{\State}{\Label}{\State'} \in \Transitions_{\Automaton_i} } \cup\mbox{} \\ 
  & \Set{
    \Filter'(\Transition)  =  \sum\limits_{\Transition' = \FromLabelTo{\Tuple{\State_L, \State}}{\Label}{\Tuple{\State_L, \State'}}} \Filter'(\Transition') \SuchThat \Transition = \FromLabelTo{\State}{\Label}{\State'} \in \Transitions_{\Automaton_j}
  }
  \end{aligned}
$$

\iffalse
Note that this definition implies that $\Filter(\Transition) =
\Filter(\Transition')$ whenever two transitions $\Transition \in
\Transitions_{\Automaton_1}, \Transition' \in \Transitions_{\Automaton_2}$
produces a product transition $\Transition'' \in \Transitions_{\Automaton_1
\times \Automaton_2}$. This corresponds to our intuition that the terms of the
product must agree on the value they accept. As before, we implicitly map
$\Filter$ to fresh terms for each transition in the product.
\fi

For instances of precisely one automaton, neither rule applies and we
perform the calculus as before.

\subsection{An Example}\label{sec:multiple:example}

We return again to our example in \cref{sec:introduction:parikh}, where we
compute the whole Parikh image of the product of $\AcaOrBc{}$ and
$\SomethingCSomething$ under the constraint that there are more instances of
letters a than b. This time our monoid $\Monoid$ is 3-dimensional vectors with
element-wise addition, and $\Map$ that maps each transition to the corresponding
increment vector, e.g $\Map(\FromLabelTo{S}{a}{A}) = \VectorLiteral{1,0,0}$.

In the interest of space, we refer back to
\cref{fig:example:single:equivalences} for the definitions of $\Filter$ for
$\AcaOrBc$ and only define $\Filter$ for $\SomethingCSomething{}$ after
substitutions as follows. Note the expansion of the $\Sigma$ labels:
    \begin{equation*}
      \begin{aligned}
        % & \Filter(\FromLabelTo{S}{a}{A}) & = 1 - \TransitionVar_6 \\
        % & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_4 + \TransitionVar_6 \\
        % & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        % & \Filter(\FromLabelTo{B}{b}{S}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        % & \Filter(\FromLabelTo{A}{a}{F}) & = 1 - \TransitionVar_6 \\
        % & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
        % Other automaton
        \Filter(\FromLabelTo{S}{\Sigma}{S}) & = \VectorLiteral{\TransitionVar_{7a}, \TransitionVar_{7b}, \TransitionVar_{7c}} \\
        \Filter(\FromLabelTo{S}{c}{F}) & = 1 \\
        \Filter(\FromLabelTo{F}{\Sigma}{F}) & = \VectorLiteral{\TransitionVar_{9a}, \TransitionVar_{9b}, \TransitionVar_{9c}} \\
      \end{aligned}
    \end{equation*}

In the derivation tree of \cref{fig:derivation:multi} we start as before, but
with the product version of the $\Image$ predicate. The only possible rule here
is $\ExpandM$, which we use to add the corresponding constraints on each
automata of the product as we would have had in the single-automaton version. We
perform algebraic reasoning on those equations, along with the constraint on
$\MonoidElement$ to determine bounds on transition variables that will enable us
to \Subsume{} and remove one of the $\Connected{}$ predicates. When we have
finished doing so, we use \Materialise{} to compute the product of the now
filtered automata. Further algebraic reasoning allows us to remove the final
instance of $\Connected$. We then perform \Expand{} to get rid of the final
$\Image$ instance, and are immediately able to $\Subsume$ the resulting
$\Connected{}$ predicate. This, again, leaves us with a set of linear
inequalities at the leaf that we can use to obtain a model of our final
$\MonoidElement$ values (now a vector). This time we can read out aca, or
$\VectorLiteral{2, 0, 1}$.

\begin{figure}
  \centering
\begin{prooftree}
  \hypo{
    {\begin{array}{lll}
      1 = \TransitionVar_{10} & \land \TransitionVar_3 = \TransitionVar_{11} & \land 1 = \TransitionVar{12} \\ 
      \land 1 = \TransitionVar_{11} &  & \land 2 > 0
    \end{array}}  
  \infer1[\Subsume{}, \Expand{}, \Subsume{}]{
    {\begin{array}{ll}
      2 > 0 \land 1 = \TransitionVar_{10} & \land \TransitionVar_3 = \TransitionVar_{11} \\
      \land 1 = \TransitionVar_{12} & \land 1 = \TransitionVar_{11} \\
      \land \Image{}_{\Automaton', \Map}(\Filter, \VectorLiteral{2, 0, 1}) & \land \Connected(\SomethingCSomething{}, \Filter) \\
    \end{array}}  
  \infer1[\EquationReasoning{}]{
  {  \begin{array}{ll}
      1 = \TransitionVar_{10} \land \TransitionVar_3 = \TransitionVar_{11} & 
      \land 1 = \TransitionVar_{12} \land 1 = \TransitionVar_{11} \\
      \land \Image{}_{\Automaton', \Map}(\Filter, 
      \VectorLiteral{
        \TransitionVar_{7a} + \TransitionVar_{9a},
        0,
        \TransitionVar_{7c} + \TransitionVar_{9c} + 1}) & \land 
        \TransitionVar_6 + \TransitionVar_4 = 0 \\
        \land 1 - \TransitionVar_6 > 0 \land \TransitionVar_6 = 0 &
        \land \TransitionVar_4 = 0 \land  \TransitionVar_3 > 0 \\
        \land 2\TransitionVar_6 + \TransitionVar_4 = 0 & \land 
          \VectorLiteral{
        \TransitionVar_{7a} + \TransitionVar_{9a},
        0,
        \TransitionVar_{7c} + \TransitionVar_{9c} + 1} 
        = \VectorLiteral{
          2,
          0,
          \TransitionVar_3} \\ \land 
      \Connected(\SomethingCSomething{}, \Filter) &
      \land \TransitionVar_{7a} + \TransitionVar_{9a} > 0
    \end{array} } 
  \infer1[\Materialise]{
    {\begin{array}{ll}
      \TransitionVar_6 + \TransitionVar_4 = 0 \land 1 - \TransitionVar_6 > 0 &
      \land \TransitionVar_6 = 0  \land \TransitionVar_4 = 0 \\
      \land \TransitionVar_3 > 0 & \land 2\TransitionVar_6 + \TransitionVar_4 = 0 \\
      \land 
      \VectorLiteral{
        \TransitionVar_{7a} + \TransitionVar_{9a},
        0,
        \TransitionVar_{7c} + \TransitionVar_{9c} + 1}
        = \VectorLiteral{
          2,
          0,
          \TransitionVar_3} & \land
      \Connected(\SomethingCSomething{}, \Filter) \\
      \land 
      \Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, 
      \VectorLiteral{\TransitionVar_{7a} + \TransitionVar_{9a}, 0, \TransitionVar_{7c} + \TransitionVar_{9c} + 1}) & \land \TransitionVar_{7a} + \TransitionVar_{9a} > 0
    \end{array}}
  \infer1[\Subsume{}]{
  {\begin{array}{ll}
    \TransitionVar_6 + \TransitionVar_4 = 0 \land 
    1 - \TransitionVar_6 > 0 &
    \land
    \TransitionVar_6 = 0 \land 
    \TransitionVar_4 = 0 \\
    \land
    \TransitionVar_3 > 0 & \land
    2\TransitionVar_6 + \TransitionVar_4 = 0 \\ 
    \land
    \VectorLiteral{
      \TransitionVar_{7a} + \TransitionVar_{9a},
      0,
      \TransitionVar_{7c} + \TransitionVar_{9c} + 1}
      = \VectorLiteral{
        2,
        0,
        \TransitionVar_3} & \land 
    \Connected(\AcaOrBc{}, \Filter) \\
    \land 
    \Connected(\SomethingCSomething{}, \Filter) \\
    \land 
    \Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, 
    \VectorLiteral{\TransitionVar_{7a} + \TransitionVar_{9a}, 0, \TransitionVar_{7c} + \TransitionVar_{9c} + 1}) & \land \TransitionVar_{7a} + \TransitionVar_{9a} > 0
  \end{array}}
  \infer1[\EquationReasoning]{
    \begin{aligned}
      \VectorLiteral{a, b, c} = \VectorLiteral{
          2 - 2\TransitionVar_6,
          2\TransitionVar_4 + 3\TransitionVar_6,
          \TransitionVar_3 + \TransitionVar_6
        } \land
        \\
        \VectorLiteral{a, b, c} = \VectorLiteral{
            \TransitionVar_{7a} + \TransitionVar_{9a},
            \TransitionVar_{7b} + \TransitionVar_{9b},
            \TransitionVar_{7c} + \TransitionVar_{9c} + 1
          } \land \\
      \Connected(\AcaOrBc{}, \Filter) \land 
      \Connected(\SomethingCSomething{}, \Filter) \land \\
      \Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, 
      \VectorLiteral{a, b, c}) \land a > b
    \end{aligned}
  \infer1[\ExpandM]{\Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, \VectorLiteral{a, b, c}) \land a > b}
\end{prooftree}
\caption{A derivation for \Calculus{} on the Parikh image of strings with more a's than b's in the product of $\AcaOrBc{}$ and $\SomethingCSomething{}$.}\label{fig:derivation:multi}
\end{figure}

\subsection{Correctness of \Calculus{} for Products of Automata}

Since \cref{tbl:rules:multi} only extends the existing rules of
\cref{tbl:rules:single}, we focus on the differences compared
to the calculus for a single automaton.

%\subsubsection{\Calculus{} for products of automata terminates}
\begin{lemma}\label{lma:multi-terminates}
  Suppose $\SomeClause{}$ is a set of formulas in which the product
  version of $\Image$ only occurs positively. There is no
  infinite sequence of proofs~$P_0, P_1, P_2, \ldots$ in which $P_0$
  has $\SomeClause{}$ as root, and each $P_{i+1}$ is derived from
  $P_i$ by applying one of the rules in \cref{tbl:rules:multi}.
\end{lemma}

\begin{proof}
  The rule \Materialise{} can similarly only be used finitely many times, as
  each application reduces the number of automata in the product of $\Image$ by
  one automaton, until only one remains and \cref{lma:single-terminates} for
  single-automaton instances applies.

  This implies that also the rule~\ExpandM{} can only be applied
  finitely often since its side condition only allows applying it once
  per instance of an $\Image$ predicate containing a product, and only finitely
  many instances of $\Image$ can be introduced on each branch.
\end{proof}

%\subsubsection{The rules in \cref{tbl:rules:multi} are solution-preserving}

Since our calculus now includes a rule introducing new variables, the
\Materialise{} rule, we have to slightly generalise the notion of
solution-preservation:
%
\begin{lemma}\label{lma:multi-correct}
  Consider an application of one of the rules in
  \cref{tbl:rules:multi}, with
  premises~$\SomeClause_1, \ldots, \SomeClause_k$ and
  conclusion~$\SomeClause$. An assignment~$\beta$ (over the symbols in
  $\SomeClause$) satisfies the conclusion~$\SomeClause$ if and only if
  there is an extension~$\beta' of \beta$ satisfying one of the
  premises~$\SomeClause_i$.
\end{lemma}

\begin{proof}
  We have to consider the two new rules in \cref{tbl:rules:multi}. The
  result is immediate for \ExpandM{}, since this rule does not remove the
  $\Image$ predicate from a proof goal, and the newly introduced formulas
  are all implied by the $\Image$ predicate.

  For \Materialise{}, observe that the existence of an accepting path
  in $\Automaton_i \times \Automaton_j$ is equivalent to the existence
  of individual paths in $\Automaton_i, \Automaton_j$ accepting the
  same word. The path in the product will satisfy the flow equations
  and connectedness, and it will be related to the individual paths as
  stipulated by the \BindingSum{} predicate.
\end{proof}

>>>>

\section{Parikh Images from Products of Automata}\label{sec:multiple}
\input{multiple.tex}
